
\ifthenelse{\boolean{short}}{
In future work, we plan to continue a preliminary investigation of
equational adequacy that is discussed in the extended version,
investigating whether that the logical adequacy proofs are an
isomorphism on $\beta\eta$-classes of derivations.  It is generally easy
to show that object-language equations are true in the framework.  We
conjecture that the converse is true for the mode theories we have
described here, which says that the ``extra'' types and judgements
available in the framework do not add to the equations between terms in
the image of encoded sequents.  
Proving this is challenging because the
equational theory of Section~\ref{sec:equational} does not itself
obviously have the subformula property---we can in principle prove
equations by introducing and then eliminating cuts.  We have sketched a
proof of equational adequacy for the simplest case (ordered logic
products), assuming a lemma that the equational theory from
Section~\ref{sec:equational} can be characterized as some ``permuting
conversions'' on cut-free derivations (i.e. that one can first
$\beta$-reduce and then rearrange the normal form).  We proved this
lemma for one-variable sequents~\citep{ls16adjoint}, and have checked
one direction here.

}{
In future work, we plan to continue the preliminary investigation of
equational adequacy that is discussed
Section~\ref{sec:adequacy-equational}.  
}
Additionally, we plan to apply our framework to investigate more
extensions of homotopy type theory like the spatial type theory
considered here; in current work with Eric Finster, we are designing a
variant of cohesion that is an internal language for spectra.  We also
plan to consider encodings of programming-focused type theories, such as
specialized effect calculi.  Finally, our adequacy proofs require
reasoning about the 1- and 2-cells in the mode theory, which we have
currently done entirely na\"ively; we would like to investigate using
techniques from higher-dimensional rewriting to simplify and possibly
automate these proofs.

